%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgements

I would like to thank \textit{Faculdade de Ciências e Tecnologia da Universidade Nova de Lisboa} and in special to the \textit{Departamento de Informática} for providing me an excelent knowledge base and for setting me challenges that would make me overcome myself, day after day.

Secondly, I need to thank my advisers Tiago Simões, Teresa Romão, and Rui Nóbrega for mentoring me throughout this dissertation. I learned a lot from you!  

Thank you to OutSystems for taking the initiative of this interesting research project. And also to all the people I met in the office and in the online meetings, due to this new normal. Everyone is ready to help and the environment is amazing. I would like to thank also all participants of the usability tests.

Furthermore, I need to give a huge thank you to my friend and partner of this adventure Pedro Deodato. We met each other in the first year of university and then we have been walking this journey as a team - P\&PCl ever. Thank you for sharing with me the greatest achievements and all the less good moments we had to overcome during these five years. We will never forget the sleepless nights working on our projects, but the most important is that we did it!

Last but not least, I would also like to express that I'm thankful for all the support my family gave since I decided to go to this course. In particular, I would like to thank you, my parents, and my siblings. They always encouraged me and supporting me whenever I needed it. Without your support, it would have been impossible to get this far. Also, a big thank you to my grandparents, you are an inspiration to me, this is for you. 


Finally, a special thank you to Mariana who always knew what to say in the good and worst moments. I am grateful to share my life with you. You are very special and an inspiration to me. This accomplishment would not have been possible without you.

