%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% chapter2.tex
%% NOVA thesis document file
%%
%% Chapter with the template manual
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Related Work}
\label{cha:related_work}
On this thesis, it is pretended to apply Human-computer interaction (HCI), Data 
Visualization and Visual Querying concepts, techniques and technologies to improve 
a Visual Querying Feature of the OutSystems Low-code Development Platform. Thus, 
in this chapter, will be presented the results of a study that analysed what is 
the Low-code development platform background and its actual situation, as well as 
what are the techniques and technologies which already exist, including some 
comparison between them. Finally, will be enumerated what products, technologies 
and tools exist on other commercial applications which can be related with the 
topics of this thesis.

\section{OutSystems Background}
\label{sec:outsystems_background}
The entirety of this thesis has the aim of improving the OutSystems Platform, so it is 
very important to understand what is that product and what can be developed with 
it.
This section provides an overview of this Low-code Development Platform, 
describing its value proposal, and its technological goals and approaches. In 
addition, it will be used some images to illustrate some relevant aspects of the 
platform.

\section{Techniques}
\label{sec:techniques}
After all the research is done, in this section will be presented a summary of 
interesting techniques found, which can be very useful to the analysis and 
development phases. The main focus is to find a set of techniques that help 
to find solutions to the problems of this thesis.

\section{Technologies}
\label{sec:technologies}
Such as the techniques research, it is also very important to search what are 
the technologies related with the subjects of this thesis that already exist, 
as this knowledge can be very important to the concept of a solution proposal.
In this way, in this section will be enumerated with a brief description of what 
are the related technologies which are relevant to this project. 

\section{Commercial Applications}
\label{sec:commercial_applications}
In any research, the academic content should not be the only taken into 
account, because sometimes the knowledge does not evolve only in the research 
centres but also in the companies. Since this thesis is made to improve a 
company product, the latter assertion has additional strength, so will be 
introduced commercial applications which can be useful to all the entirety of 
this process.